Object-oriented implementations of the MPDATA advection equation solver in~C++,~Python~and~Fortran

### 2013-03-18 paper/doc/paper.tex
further text cleanups (with missing paper.tex)

% for conpact itemize etc
\usepackage{enumitem}
\setitemize{noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt}

@@

\journal{Computer Physics Communications}
%\journal{arXiv. Paper presented at the American Meteorological Society 2013 Annual Meeting in Austin, Texas, USA}

@@

    \begin{abstract}
        Three object-oriented implementations of a prototype solver of the advection equation are introduced.
        Presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
        The solvers include an implementation of the Multidimensional Positive-Definite 
          Advective Transport Algorithm (MPDATA).
        The introduced codes exemplify how the application of 
          object-oriented programming (OOP) techniques allows to reproduce the mathematical notation 
          used in the literature within the program code.
        The codes serve also as a basis for discussion on the tradeoffs of the programming language choice.
        The main angles of comparison are code brevity and syntax clarity
          (and hence maintainability and auditability) as well as performance.
        In case of Python, a significant performance gain is observed when switching from the standard 
          interpreter (CPython) to the PyPy implementation of Python.
        Entire source code of all three implementations is embedded in the text and is licensed
          under the terms of the GNU GPL license. 
    \end{abstract}

@@

  Object oriented programming (OOP) {\em ''has become recognised as the almost unique successful 
    paradigm for creating complex software''} \citep[][Sec.~1.3]{Press_et_al_2007}.
  It is intriguing that while the quoted statement comes from the very book subtitled 
   {\em The Art of Scientific Computing}, hardly any (if not none) of currently operational 
    weather and climate prediction systems - flagship examples of complex scientific software - 
    makes extensive use of OOP techniques\footnote{
  Fortran have been the language of choice in oceanic \citep{Griffies_et_al_2000}, 
    weather-prediction \citep{Sundberg_2009} and Earth system \citep{Legutke_2012} modelling, 
    and none of its 20-century editions were object-oriented languages \citep[see e.g.][for discussion]{Norton_et_al_2007}.}.
  Application of OOP techniques in development of numerical modelling software may help to
    {\bf (i)} maintain modularity and separation of program logic layers (e.g. separation of
      a dynamical core, parallelisation mechanisms, data input/output, error handling and
      the description of physical processes); and
    {\bf (ii)} {\bf shorten and simplify the source code and improve its readability by reproducing within 
      the program logic the mathematical abstractions and notation used in the literature}.
  The first application is attainable, yet arguably cumbersome, with procedural programming.
  The latter, virtually impossible to obtain with procedural programming, is the focus of this paper.
  It also enables the compiler or library authors to relieve the user (i.e. scientific programmer)
    from hand-coding (premature) optimisations, a practice long recognised as having {\em a strong negative impact when debugging
    and maintenance are considered} \citep{Knuth_1974}.

  MPDATA \citep{Smolarkiewicz_1984} stands for Multidimensional Positive Definite Advective Transport Algorithm and is
    an example of a numerical procedure used in weather, climate and ocean simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002}.
  MPDATA is a solver for systems of equations of the following form:
  \begin{equation}\label{eq:adv}
    \input{formulae/adv}
  \end{equation}
  that describe transport of quantity $\psi$ by fluid flow with velocity $\vec{v}$.
  Quoting Numerical Recipes once more, development of methods to numerically solve such problems 
    {\em ''is an art as much as a science''} \citep[][Sec.~20.1]{Press_et_al_2007},
    and MPDATA is an example of the state-of-the art in this field.
  MPDATA is designed to accurately solve equation (\ref{eq:adv}) in arbitrary
    number of dimensions assuring positive-definiteness of scalar field $\psi$ 
    and incurring small numerical diffusion.
  All relevant MPDATA formul\ae are given in the text but are presented without
    derivation or detailed discussion.
  For a recent review of MPDATA-based techniques see \citet[][and references therein]{Smolarkiewicz_2006}.

  In this paper we introduce and discuss object-oriented implementations of an MPDATA-based 
    two-dimensional (2D) advection equation solver written in C++11 (\citeauthor{ISO_14882} 14882:2011), 
    Python \citep{Rossum_and_Drake_2011} and Fortran 2008 (\citeauthor{ISO_1539-1} 1539-1:2010). 
  In the following section we introduce the three implementations
    briefly describing the algorithm itself and
    discussing where and how the OOP techniques may be applied in its implementation.
  The syntax and nomenclature of OOP techniques are used without introduction,
    for an overview of OOP in context of C++, Python and Fortran, consult for example
    \citep[][Part~II]{Stroustrup_2000}, \citep[][Chapter~5]{Pilgrim_2004} and
    \citep[][Chapter~11]{Markus_2012}, respectively.
  The third section covers performance evaluation of the three implementations.
  The fourth section covers discussion of the tradeoffs of the programming language choice.
  The fifth section closes the article with a brief summary.

  Throughout the paper the three implementations are presented by discussing 
    source code listings which cover entire program code.
  Subsections \ref{sec:array}-\ref{sec:solver} describe all three implementations,
    while subsequent sections \ref{sec:cyclic}-\ref{sec:example} cover discussion of C++ code 
    only.
  The relevant parts of Python and Fortran codes do not differ significantly, and for readability reasons 
    are presented in the \ref{app:P} and the \ref{app:F}, respectively.

@@

  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation onto a grid of the $\psi$ and 
    the Courant number $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ fields, where $\Delta t$ is the 
    solver timestep and $\Delta x$ is the grid spacing.
  Among the three considered languages only Fortran is equipped with built-in
    array handling facilities of practical use in high-performance computing.

  Presented C++ implementation of MPDATA is built upon the Blitz++ 
    library\footnote{Blitz++ is a C++ class library for scientific computing which uses the expression templates technique to achieve high performance, see http://sf.net/projects/blitz/}.
  Blitz offers object-oriented representation of n-dimensional arrays,
    and array-valued mathematical expressions.
  In particular, it offers loop-free notation for array arithmetics
    that does not incur creation of intermediate temporary objects.
  Blitz++ is a header-only library\footnote{Blitz++ requires linking with \prog{libblitz} if debugging mode is used} -- to use it, it is enough to include the appropriate header file,
    and optionally expose the required classes to the present namespace:
  \codecpp{code/cpp/listings.hpp}{listing02}{listing03}{1}
  Here \prog{arr\_t}, \prog{rng\_t} and \prog{idx\_t} serve as alias identifiers 
    and are introduced in order to shorten the code.

  Arguably, the power of Blitz++ comes from the ability to express array expressions as objects.
  In particular, it is possible to define a function that returns an array expression;
    i.e. not the resultant array, but an object representing a ,,recipe'' defining the operations
    to be performed on the arguments.
  As a consequence, the return types of such functions become unintelligible.
  Luckily, the \prog{auto} return type declaration from the C++11 standard allows to simplify the code significantly,
    even more if used through the following preprocessor macro:
  \codecpp{code/cpp/listings.hpp}{listing03}{listing04}{1}
  The call to \prog{blitz::safeToReturn()} function is included in order to ensure that
    all arrays involved in the expression being returned continue to exist in the
    caller scope.
  For example, definition of a function returning its array-valued argument doubled reads:
    \prog{auto~f(arr\_t~x)~return\_macro(2*x)}.
  This is the only preprocessor macro defined herein.

  For the Python implementation of MPDATA the NumPy package is used.
  In order to make the code compatible with both the standard CPython
    as well as the alternative PyPy implementation of Python \citep[][]{Bolz_et_al_2011},
    the Python implementation includes the following sequence of \prog{import} statements:
  \codepyt{code/pyt/listings.py}{listing02}{listing03}{1}
  First, the PyPy's built-in NumPy implementation named \prog{numpypy} is imported if applicable (i.e. if running PyPy), 
    and the lazy evaluation mode is turned on through the \prog{set\_invalidation(False)} call.
  PyPy's lazy evaluation obtained with the help of a just-in-time compiler enables to achieve
    an analogous to Blitz++ temporary-array-free handling of array-valued expressions 
    (see discussion in section~\ref{sec:perf}).
  Second, to match the settings of C++ and Fortran compilers used herein, the NumPy package is instructed 
    to ignore any floating-point errors (only if such an option
    is available in the interpreter\footnote{\prog{numpy.seterr()} is not supported in PyPy as of version 1.9}).
  The above lines conclude all code modifications that needed to be added in order to run
    the code with PyPy.

  Fortran features built-in multidimensional array handling mechanisms,
    hence there is no need for using an external package as with C++ and Python.
  Fortran array-handling features are not object-oriented, though.

  \subsection{Array sequence containers}\label{sec:sequence}

  While discretisation in space of the scalar field $\psi(x,y)$ into its $\psi_{[i,j]}$ 
    grid representation requires floating-point array containers (cf. section \ref{sec:array}).
  Discretisation in time requires a container class for storing
    sequences of such arrays, i.e. \{$\psi^{[n]}$, $\psi^{[n+1]}$\}.
  Similarly the components of the vector field $\vec{C}$ are in fact a \{$C^{[x]}$, $C^{[y]}$\} 
    array sequence.

  Using an additional array dimension to represent the sequence elements is not considered for two reasons.
  First, the $C^{[x]}$ and $C^{[y]}$ arrays constituting the sequence have different sizes
    (see discussion of the Arakawa-C grid in section~\ref{sec:grid}).
  Second, the order of dimensions would need to be different for different languages to assure that
      the contiguous dimension is not used for time levels.

  In the C++ implementation the Boost\footnote{
    Boost is a free and open-source collection of peer-reviewed C++ libraries available at \url{http://www.boost.org/}.
    Several parts of Boost have been integrated into or inspired new additions to the C++ standard.
  } \prog{ptr\_vector} class is used to represent sequences of Blitz++ arrays 
    and at the same time to handle automatic freeing of dynamically allocated memory.
  For clearer ,,translation'' of mathematical formul\ae~into C++ code, the
    \prog{ptr\_vector} class is further customised by defining a derived structure whose element-access \prog{[~]} 
    operator is overloaded with a modulo-variant:
  \codecpp{code/cpp/listings.hpp}{listing04}{listing05}{1}
  Consequently the last element of any such sequence may be accessed at index \prog{-1}, the last but one at \prog{-2}, 
    and so on.

@@

  Fortran does not feature any built-in sequence container capable of storing arrays, 
    hence a custom \prog{arrvec\_t} type is introduced:
  \codefor{code/for/listings.f}{listing02}{listing03}{1}
  The \prog{arr\_t} type is defined solely for the purpose 
    of overcoming the Fortran's limitation of not allowing to declare arrays of arrays.
  The \prog{arrptr\_t} type is defined as creating an allocatable of its instances instead
    of multi-element pointer of \prog{arr\_t} ensures automatic memory deallocation.
  The \prog{ctor()} method initialises the container for a given number of elements \prog{n}.
  The \prog{init()} method initialises the \prog{n}-th element of the container with 
    an newly allocated 2D array spanning indices \prog{i(1)}:\prog{i(2)}, 
    and \prog{j(1)}:\prog{j(2)} in the first, and last dimensions respectively\footnote{In Fortran, when an array
    is passed as a function argument its base is locally set to unity, regardless of the setting
    at the caller scope.}.
  Access to the arrays stored in the container is provided through the \prog{at} member field.
  Assuming \prog{psi} is an instance of \prog{arrptr\_t}, the \prog{(i,j)} element of the \prog{n}-th array in \prog{psi}
    may be accessed with \prog{psi\%at( n )\%p\%a( i, j )}.

@@

    \draw [ultra thick,-latex,red] (0,.8) -- (0,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (0,-1.2) -- (0,-.8) node [below right] {$C^{[y]}_{[i,j\mhlf]}$};
    \draw [ultra thick,-latex,red] (-2,.8) -- (-2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (-2,-1.2) -- (-2,-.8) node [above] {};
    \draw [ultra thick,-latex,red] (2,.8) -- (2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (2,-1.2) -- (2,-.8) node [above] {};
  \end{tikzpicture}
  \caption{\label{fig:grid}
    A schematic of the Arakawa-C grid.
  }
  \end{figure}
  The so-called Arakawa-C staggered grid \citep{Arakawa_and_Lamb_1977} depicted in Figure~\ref{fig:grid}
    is a natural choice for MPDATA.
%  MPDATA is a finite-difference approximation to the flux-form of the equation~(\ref{eq:adv}) 
%    resulting in a finite-volume scheme.
%  Using the Arakawa-C grid allows to evaluate the fluxes at 
%    the grid cell (finite volume) edges without interpolation.
  As a consequence, the discretised representations of the $\psi$ scalar field, and each 
    component of the ${\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}}$ vector field 
    in eq.~(\ref{eq:adv}) are defined over different grid point locations.
  In mathematical notation this can be indicated by usage of fractional indices, e.g.
    $C^{[x]}_{[i\mhlf,j]}$, $C^{[x]}_{[i+\phlf,j]}$, $C^{[y]}_{[i,j\mhlf]}$ and $C^{[y]}_{[i,j+\phlf]}$
    to depict the grid values of the ${\vec{C}}$ vector components surrounding $\psi_{[i,j]}$.
  However, fractional indexing does not have a built-in counterpart in any of the
    employed programming languages.
  A desired syntax would translate:
  \begin{itemize}
    \item{\prog{$i-\phlf$} ~~ $\leadsto$ ~~ \prog{$i-1$}} 
    \item{\prog{$i+\phlf$} ~~ $\leadsto$ ~~ \prog{$i$}.}
  \end{itemize}
  OOP offers a convenient way to implement such notation
    by overloading the \prog{+} and \prog{-} operators for objects representing array indices.

@@

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{Shift} class with the adequate operator overloads:
  \codepyt{code/pyt/listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}
    \footnote{\label{fnt:slice}One could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate class}:
  \codepyt{code/pyt/listings.py}{listing04}{listing05}{1}

@@

  \subsection{Array index permutations}\label{sec:pi}
  Hereinafter, the $\pi_{a,b}^{d}$ symbol is used to denote a  
    cyclic permutation of the order $d$ of the set $\{a,b\}$.
  It is used to generalise the MPDATA formul\ae~into multiple dimensions
    using the following notation:
  \begin{equation*}\label{eq:pi}
    \input{formulae/pi}
  \end{equation*}

@@

  In the Fortran implementation \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code/for/listings.f}{listing05}{listing06}{1}
  \prog{span()} is used to shorten the declarations of arrays to be returned 
    from functions in the Fortran implementation (see listings F.11 and F.17--F.20).

@@

  \subsection{A prototype solver}\label{sec:solver}

  The tasks to be handled by a prototype advection equation solver proposed herein are:
  \begin{description}
    \item[(i)]{storing arrays representing the $\psi$ and $\vec{C}$ fields and any required housekeeping data,}
    \item[(ii)]{allocating and deallocating the required memory,}
    \item[(iii)]{providing access to the solver state, and}
    \item[(iv)]{performing the integration by invoking the advection-operator and
      boundary-condition handling routines.}
  \end{description}
  In the following C++ definition of the \prog{solver} structure, task (i) is represented with the definition of the structure
    member fields; task (ii) is split between the \prog{solver}'s constructor and the destructors of \prog{arrvec\_t};
    task (iii) is handled by the accessor methods; task (iv) is handled within the \prog{solve} method:
  \codecpp{code/cpp/listings.hpp}{listing08}{listing09}{1}
  The \prog{solver} structure is an abstract definition (containing a pure virtual method) 
     requiring its descendants to implement at least
     the \prog{advop()} method which is expected to fill \prog{psi[n+1]} with an updated (advected)
     values of \prog{psi[n]}.
  The two template parameters \prog{bcx\_t} and \prog{bcy\_t} allow the solver to operate with 
    any kind of boundary condition structures that fulfil the requirements implied by the calls to the 
    methods of \prog{bcx} and \prog{bcy}, respectively.

  The donor-cell and MPDATA schemes both require only the previous state of an advected field
    in order to advance the solution.
  Consequently, memory for two time levels ($\psi^{[n]}$ and $\psi^{[n+1]}$) is allocated in the
    constructor.
  The sizes of the arrays representing the two time-levels of $\psi$ 
    are defined by the domain size ({\em nx}$~\times$~{\em ny}) plus the halo region.
  The size of the halo region is an argument of the constructor.
  The \prog{cycle()} method is used to swap the time levels without copying any data.

>>>>

% for conpact itemize etc
\usepackage{enumitem}
\setitemize{noitemsep,itemsep=1pt,topsep=1pt,parsep=0pt,partopsep=0pt}

@@

\journal{Computer Physics Communications}

@@

    \begin{abstract}
        Three object-oriented implementations of a prototype solver of the advection equation are introduced.
        The presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
        The solvers include an implementation of the Multidimensional Positive-Definite 
          Advective Transport Algorithm (MPDATA).
        The introduced codes exemplify how the application of 
          object-oriented programming (OOP) techniques allows to reproduce the mathematical notation 
          used in the literature within the program code.
        A discussion on the tradeoffs of the programming language choice is presented.
        The main angles of comparison are code brevity and syntax clarity
          (and hence maintainability and auditability) as well as performance.
        In the case of Python, a significant performance gain is observed when switching from the standard 
          interpreter (CPython) to the PyPy implementation of Python.
        Entire source code of all three implementations is embedded in the text and is licensed
          under the terms of the GNU GPL license. 
    \end{abstract}

@@

  Object oriented programming (OOP) {\em ''has become recognised as the almost unique successful 
    paradigm for creating complex software''} \citep[][Sec.~1.3]{Press_et_al_2007}.
  It is intriguing that, while the quoted statement comes from the very book subtitled 
   {\em The Art of Scientific Computing}, hardly any (if not none) of the currently operational 
    weather and climate prediction systems - flagship examples of complex scientific software - 
    make extensive use of OOP techniques.
  Fortran has been the language of choice in oceanic \citep{Griffies_et_al_2000}, 
    weather-prediction \citep{Sundberg_2009} and Earth system \citep{Legutke_2012} modelling, 
    and none of its 20-century editions were object-oriented languages \citep[see e.g.][for discussion]{Norton_et_al_2007}.

  Application of OOP techniques in development of numerical modelling software may help to:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=ii]
    \item{maintain modularity and separation of program logic layers (e.g. separation of
      numerical algorithms, parallelisation mechanisms, data input/output, error handling and
      the description of physical processes); and}
    \item{{\bf shorten and simplify the source code and improve its readability by reproducing within 
      the program logic the mathematical notation used in the literature}.}
  \end{enumerate}
  The first application is attainable, yet arguably cumbersome, with procedural programming.
  The latter, virtually impossible to obtain with procedural programming, is the focus of this paper.
  It also enables the compiler or library authors to relieve the user (i.e. scientific programmer)
    from hand-coding optimisations, a practice long recognised as having {\em a strong negative impact when debugging
    and maintenance are considered} \citep{Knuth_1974}.

  MPDATA \citep{Smolarkiewicz_1984} stands for Multidimensional Positive Definite Advective Transport Algorithm and is
    an example of a numerical procedure used in weather, climate and ocean simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002}.
  MPDATA is a solver for systems of advection equations of the following form:
  \begin{equation}\label{eq:adv}
    \input{formulae/adv}
  \end{equation}
  that describe evolution of a scalar field $\psi$ transported by the fluid flow with velocity $\vec{v}$.
  Quoting Numerical Recipes once more, development of methods to numerically solve such problems 
    {\em ''is an art as much as a science''} \citep[][Sec.~20.1]{Press_et_al_2007},
    and MPDATA is an example of the state-of-the art in this field.
  MPDATA is designed to accurately solve equation (\ref{eq:adv}) in an arbitrary
    number of dimensions assuring positive-definiteness of scalar field $\psi$ 
    and incurring small numerical diffusion.
  All relevant MPDATA formul\ae~are given in the text but are presented without
    derivation or detailed discussion.
  For a recent review of MPDATA-based techniques see \citet[][and references therein]{Smolarkiewicz_2006}.

  In this paper we introduce and discuss object-oriented implementations of an MPDATA-based 
    two-dimensional (2D) advection equation solver written in C++11 (\citeauthor{ISO_14882} 14882:2011), 
    Python \citep{Rossum_and_Drake_2011} and Fortran 2008 (\citeauthor{ISO_1539-1} 1539-1:2010). 
  In the following section we introduce the three implementations
    briefly describing the algorithm itself and
    discussing where and how the OOP techniques may be applied in its implementation.
  The syntax and nomenclature of OOP techniques are used without introduction,
    for an overview of OOP in context of C++, Python and Fortran, consult for example
    \citep[][Part~II]{Stroustrup_2000}, \citep[][Chapter~5]{Pilgrim_2004} and
    \citep[][Chapter~11]{Markus_2012}, respectively.
  The third section of this paper covers performance evaluation of the three implementations.
  The fourth section covers discussion of the tradeoffs of the programming language choice.
  The fifth section closes the article with a brief summary.

  Throughout the paper we present the three implementations by discussing 
    source code listings which cover the entire program code.
  Subsections \ref{sec:array}-\ref{sec:solver} describe all three implementations,
    while subsequent sections \ref{sec:cyclic}-\ref{sec:example} cover discussion of C++ code 
    only.
  The relevant parts of Python and Fortran codes do not differ significantly, and for readability reasons 
    are presented in \ref{app:P} and \ref{app:F}, respectively.

@@

  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation onto a grid of the $\psi$ and 
    the Courant number $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ fields, where $\Delta t$ is the 
    solver timestep and $\Delta x$ is the grid spacing.

  Presented C++ implementation of MPDATA is built upon the Blitz++ 
    library\footnote{Blitz++ is a C++ class library for scientific computing which uses the expression templates technique to achieve high performance, see \url{http://sf.net/projects/blitz/}}.
  Blitz offers object-oriented representation of n-dimensional arrays,
    and array-valued mathematical expressions.
  In particular, it offers loop-free notation for array arithmetics
    that does not incur creation of intermediate temporary objects.
  Blitz++ is a header-only library\footnote{Blitz++ requires linking with \prog{libblitz} if debugging mode is used} -- to use it, it is enough to include the appropriate header file,
    and optionally expose the required classes to the present namespace:
  \codecpp{code/cpp/listings.hpp}{listing02}{listing03}{1}
  Here \prog{arr\_t}, \prog{rng\_t} and \prog{idx\_t} serve as alias identifiers 
    and are introduced in order to shorten the code.

  The power of Blitz++ comes from the ability to express array expressions as objects.
  In particular, it is possible to define a function that returns an array expression;
    i.e. not the resultant array, but an object representing a ,,recipe'' defining the operations
    to be performed on the arguments.
  As a consequence, the return types of such functions become unintelligible.
  Luckily, the \prog{auto} return type declaration from the C++11 standard allows to simplify the code significantly,
    even more if used through the following preprocessor macro:
  \codecpp{code/cpp/listings.hpp}{listing03}{listing04}{1}
  The call to \prog{blitz::safeToReturn()} function is included in order to ensure that
    all arrays involved in the expression being returned continue to exist in the
    caller scope.
  For example, definition of a function returning its array-valued argument doubled, reads:
    \prog{auto~f(arr\_t~x)~return\_macro(2*x)}.
  This is the only preprocessor macro defined herein.

  For the Python implementation of MPDATA the NumPy\footnote{NumPy is a Python package
    for scientific computing offering support for multi-dimensional arrays and a library
    of numerical algorithms, see \url{http://numpy.org/}} package is used.
  In order to make the code compatible with both the standard CPython
    as well as the alternative PyPy implementation of Python \citep[][]{Bolz_et_al_2011},
    the Python code includes the following sequence of \prog{import} statements:
  \codepyt{code/pyt/listings.py}{listing02}{listing03}{1}
  First, the PyPy's built-in NumPy implementation named \prog{numpypy} is imported if applicable (i.e. if running PyPy), 
    and the lazy evaluation mode is turned on through the \prog{set\_invalidation(False)} call.
  PyPy's lazy evaluation obtained with the help of a just-in-time compiler enables to achieve
    an analogous to Blitz++ temporary-array-free handling of array-valued expressions 
    (see discussion in section~\ref{sec:perf}).
  Second, to match the settings of C++ and Fortran compilers used herein, the NumPy package is instructed 
    to ignore any floating-point errors, if such an option
    is available in the interpreter\footnote{\prog{numpy.seterr()} is not supported in PyPy as of version 1.9}.
  The above lines conclude all code modifications that needed to be added in order to run
    the code with PyPy.

  Among the three considered languages only Fortran is equipped with built-in
    array handling facilities of practical use in high-performance computing. 
  Therefore, there is no need for using an external package as with C++ and Python.
  Fortran array-handling features are not object-oriented, though.

  \subsection{Containers for sequences of arrays}\label{sec:sequence}

  As discussed above, discretisation in space of the scalar field $\psi(x,y)$ into its $\psi_{[i,j]}$ 
    grid representation requires floating-point array containers.
  In turn, discretisation in time requires a container class for storing
    sequences of such arrays, i.e. \{$\psi^{[n]}$, $\psi^{[n+1]}$\}.
  Similarly the components of the vector field $\vec{C}$ are in fact a \{$C^{[x]}$, $C^{[y]}$\} 
    array sequence.

  Using an additional array dimension to represent the sequence elements is not considered for two reasons.
  First, the $C^{[x]}$ and $C^{[y]}$ arrays constituting the sequence have different sizes
    (see discussion of the Arakawa-C grid in section~\ref{sec:grid}).
  Second, the order of dimensions would need to be different for different languages to assure that
    the contiguous dimension is used for one of the space dimensions and not for time levels.

  In the C++ implementation the Boost\footnote{
    Boost is a free and open-source collection of peer-reviewed C++ libraries available at \url{http://www.boost.org/}.
    Several parts of Boost have been integrated into or inspired new additions to the C++ standard.
  } \prog{ptr\_vector} class is used to represent sequences of Blitz++ arrays 
    and at the same time to handle automatic freeing of dynamically allocated memory.
  The \prog{ptr\_vector} class is further customised by defining a derived structure which element-access \prog{[~]} 
    operator is overloaded with a modulo variant:
  \codecpp{code/cpp/listings.hpp}{listing04}{listing05}{1}
  Consequently the last element of any such sequence may be accessed at index \prog{-1}, the last but one at \prog{-2}, 
    and so on.

@@

  Fortran does not feature any built-in sequence container capable of storing arrays, 
    hence a custom \prog{arrvec\_t} type is introduced:
  \codefor{code/for/listings.f}{listing02}{listing03}{1}
  The \prog{arr\_t} type is defined solely for the purpose of overcoming the limitation 
    of lack of an array-of-arrays construct, and its only member field is a two-dimensional array.
  An array of \prog{arr\_t} is used hereinafter as a container for sequences of arrays.

  The \prog{arrptr\_t} type is defined solely for the purpose of overcoming Fortran's limitation
    of not supporting allocatables of pointers.
  \prog{arrptr\_t}'s single member field is a pointer to an instance of \prog{arr\_t}.
  Creating an allocatable of \prog{arrptr\_t}, instead of 
    a multi-element pointer of \prog{arr\_t}, ensures automatic memory deallocation.

  Type \prog{arrptr\_t} is used to implement the from-the-end addressing of elements in \prog{arrvec\_t}.
  The array data is stored in the \prog{arrs} member field (of type \prog{arr\_t}).
  The \prog{at} member field (of type \prog{arrptr\_t}) stores pointers to the elements of \prog{arrs}. 
  \prog{at} has double the length of \prog{arrs} and is initialised in a cyclic manner so that
    the \prog{-1} element of \prog{at} points to the last element of \prog{arrs}, and so on.
  Assuming \prog{psi} is an instance of \prog{arrptr\_t}, the \prog{(i,j)} element of the \prog{n}-th array in \prog{psi}
    may be accessed with\\ \prog{psi\%at( n )\%p\%a( i, j )}.

  The \prog{ctor(n)} method initialises the container for a given number of elements \prog{n}.
  The \prog{init(n,i,j)} method initialises the \prog{n}-th element of the container with 
    a newly allocated 2D array spanning indices \prog{i(1)}:\prog{i(2)}, 
    and \prog{j(1)}:\prog{j(2)} in the first, and last dimensions respectively\footnote{In Fortran, when an array
    is passed as a function argument its base is locally set to unity, regardless of the setting
    at the caller scope.}.

@@

    \draw [ultra thick,-latex,red] (0,.8) -- (0,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (0,-1.2) -- (0,-.8) node [below right] {$C^{[y]}_{[i,j\mhlf]}$};
    \draw [ultra thick,-latex,red] (-2,.8) -- (-2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (-2,-1.2) -- (-2,-.8) node [above] {};
    \draw [ultra thick,-latex,red] (2,.8) -- (2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (2,-1.2) -- (2,-.8) node [above] {};
  \end{tikzpicture}
  \caption{\label{fig:grid}
    A schematic of the Arakawa-C grid.
  }
  \end{figure}
  The so-called Arakawa-C staggered grid \citep{Arakawa_and_Lamb_1977} depicted in Figure~\ref{fig:grid}
    is a natural choice for MPDATA.
  As a consequence, the discretised representations of the $\psi$ scalar field, and each 
    component of the ${\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}}$ vector field 
    in eq.~(\ref{eq:adv}) are defined over different grid point locations.
  In mathematical notation this can be indicated by usage of fractional indices, e.g.
    $C^{[x]}_{[i\mhlf,j]}$, $C^{[x]}_{[i+\phlf,j]}$, $C^{[y]}_{[i,j\mhlf]}$ and $C^{[y]}_{[i,j+\phlf]}$
    to depict the grid values of the ${\vec{C}}$ vector components surrounding $\psi_{[i,j]}$.
  However, fractional indexing does not have a built-in counterpart in any of the
    employed programming languages.
  A desired syntax would translate \prog{$i-\phlf$} to \prog{$i-1$} and
    \prog{$i+\phlf$} to \prog{$i$}.
  OOP offers a convenient way to implement such notation
    by overloading the \prog{+} and \prog{-} operators for objects representing array indices.

@@

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{shift} class with the adequate operator overloads:
  \codepyt{code/pyt/listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}
    \footnote{\label{fnt:slice}One could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate object}:
  \codepyt{code/pyt/listings.py}{listing04}{listing05}{1}

@@

  \subsection{Array index permutations}\label{sec:pi}
  Hereinafter, the $\pi_{a,b}^{d}$ symbol is used to denote a  
    cyclic permutation of an order $d$ of a set $\{a,b\}$.
  It is used to generalise the MPDATA formul\ae~into multiple dimensions
    using the following notation:
  \begin{equation*}\label{eq:pi}
    \input{formulae/pi}
  \end{equation*}

@@

  In the Fortran implementation \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code/for/listings.f}{listing05}{listing06}{1}
  The \prog{span()} function is used to shorten the declarations of arrays to be returned 
    from functions in the Fortran implementation (see listings F.11 and F.17--F.20).

@@

  \subsection{Prototype solver}\label{sec:solver}

  The tasks to be handled by a prototype advection equation solver proposed herein are:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=iii]
    \item{storing arrays representing the $\psi$ and $\vec{C}$ fields and any required housekeeping data,}
    \item{allocating and deallocating the required memory,}
    \item{providing access to the solver state,}
    \item{performing the integration by invoking the advection-operator and
      boundary-condition handling routines.}
  \end{enumerate}
  In the following C++ definition of the \prog{solver} structure, task (i) is represented with the definition of the structure
    member fields; task (ii) is split between the \prog{solver}'s constructor and the destructors of \prog{arrvec\_t};
    task (iii) is handled by the accessor methods; task (iv) is handled within the \prog{solve} method:
  \codecpp{code/cpp/listings.hpp}{listing08}{listing09}{1}
  The \prog{solver} structure is an abstract definition (containing a pure virtual method) 
     requiring its descendants to implement at least
     the \prog{advop()} method which is expected to fill \prog{psi[n+1]} with an updated (advected)
     values of \prog{psi[n]}.
  The two template parameters \prog{bcx\_t} and \prog{bcy\_t} allow the solver to operate with 
    any kind of boundary condition structures that fulfil the requirements implied by the calls to the 
    methods of \prog{bcx} and \prog{bcy}, respectively.

  The donor-cell and MPDATA schemes both require only the previous state of an advected field
    in order to advance the solution.
  Consequently, memory for two time levels ($\psi^{[n]}$ and $\psi^{[n+1]}$) is allocated in the
    constructor.
  The sizes of the arrays representing the two time levels of $\psi$ 
    are defined by the domain size ({\em nx}$~\times$~{\em ny}) plus the halo region.
  The size of the halo region is an argument of the constructor.
  The \prog{cycle()} method is used to swap the time levels without copying any data.
